---
title: "Genotype phenotype sAML1"
author: 
- Zhan Yinxiu
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 6,
                      fig.height = 6)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(mclust)
  library(maftools)
  library(ggplot2)
  library(plotly)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(ReactomePA)
  source("src/functions.R")
})

```

# Objective

The objective of the analysis is to integrate the genotype data from the enrichemnt experiment with the expression data.
We start with sAML1. 

# Genotype data

The genotype data has been analyzed using [NanoGen](https://github.com/dimadatascience/NanoGen), a custom pipeline designed to call mutations from targeted enrichment data. NanoGen follows a multi-step process to ensure accurate mutation detection. It starts by using Blaze to identify cell barcodes, ensuring that each cell’s data is correctly assigned. It then collapses Unique Molecular Identifiers (UMIs) to eliminate PCR duplicates and other biases, allowing for a more accurate representation of the original sample.

One key feature of NanoGen is its ability to call consensus at the base level, rather than at the sequence level, providing a more precise identification of mutations. NanoGen applies a negative binomial model to distinguish between true mutated reads and background noise, improving the reliability of mutation calls. To address the issue of mutated allele dropout, NanoGen imposes a strict threshold: it requires at least five distinct wild-type (WT) UMIs for a cell to be confidently classified as wild-type (WT).

## Data exploration

First, we identify and retain only mutations that overlap with the WES data. For each candidate mutation, we calculate its cellular fraction. Next, we filter out any genes with fewer than 50 genotyped cells.

```{r}
genotype_file = "data/enrichment/sAML1.csv"
wes_file = "data/S29942_sAML1_D.small_mutations.cancervar.escat.maf"
sample_id = "sAML1"
```

```{r, results=FALSE}
# Read and filter genotyped cells
genotype = read.csv(genotype_file, row.names = 1)
genotype = genotype[genotype$genotype!="", ]
genotype['total_umis'] = (genotype[,'WT'] + genotype[,'MUT'] + genotype[,'MIS'])
genotype$alt <- ifelse(grepl('-', genotype$alt), '-', genotype$alt)
genotype['id'] = paste0(genotype[,'gene'], genotype[,'start'], genotype[,'alt'])

variants = data.frame(read.maf(wes_file)@data)

variants['id'] = paste0(variants[,'Hugo_Symbol'] , variants[,'Start_Position'] , variants[,'Tumor_Seq_Allele2'])
genotype = as_tibble(genotype[genotype$id %in% (variants$id), ])

# Calculate ccf and filter lowly genotyped mutations
ccf = genotype %>%
  group_by(gene, genotype) %>%            # Group by 'gene' and 'genotype'
  summarise(count = n(), .groups = 'drop') %>%  # Count occurrences of each 'genotype'
  pivot_wider(names_from = genotype,
              values_from = count,
              values_fill = 0)  # Pivot to wide format, filling missing values with 0

ccf = ccf[(ccf$MUT + ccf$WT) > 50,]
ccf['ccf'] = ccf[,'MUT'] /(ccf[,'MUT'] + ccf[,'WT']) 

# Add wes
subset <- variants %>%
  filter(id %in% genotype$id) %>%
  select(Hugo_Symbol, tumor_f)

ccf <- ccf %>%
  left_join(subset, by = c("gene" = "Hugo_Symbol"))  %>%
  mutate(wes_ccf = pmin(tumor_f * 2, 1))  # Ensure values > 1 are set to 1

# Calculate correlation
correlation <- cor(ccf$ccf, ccf$wes_ccf, use = "complete.obs")
correlation_title <- sprintf("Correlation: %.2f", correlation)



# Create the scatter plot
p <- ggplot(ccf, aes(x = ccf, y = wes_ccf, text = gene)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  ggtitle(correlation_title) +
  labs(x = "single cell CF", y = "WES_CF") +
  theme(plot.title = element_text(hjust = 0.5, size = 20))

# Convert to interactive plotly plot
fig <- ggplotly(p, tooltip = c("text"))
```

Below, we present the correlation between the cellular fraction, defined as VAF * 2, in WES  and the cellular fraction in the target enrichment. As demonstrated, the target enrichment effectively recapitulates the cellular fraction observed in WES.

```{r}
fig
```


```{r}
DT::datatable(
  ccf,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```


## Mutation co-occurence  

Next, we quantified the number of cells with multiple genotyped mutations, as shown in the tables below. We present the results for different mutation combinations. For this specific samples, we will look at these genes since they are the most genotyped: RUNX1, SRSF2, IDH1, FLT3, CSF3R

```{r}
# Define the genes of interest
gene_interest <- c('FLT3', 'IDH1', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)

gene_interest <- c('FLT3', 'SRSF2', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)


gene_interest <- c('CSF3R', 'SRSF2', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)


gene_interest <- c('CSF3R', 'IDH1', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)

gene_interest <- c('FLT3', 'SRSF2','IDH1', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```


# Genotype phenotype analysis: full genotype

We begin by analyzing the phenotype-genotype relationships using only cells with genotyped data for multiple genes. Specifically, we will focus on the following cases:

1- RUNX1 - IDH1 - FLT3

2- RUNX1 - SRSF2 - FLT3

The reason is that, for these combination, we have the highest number of genotyped cells for all genes.

For each case, we will examine the impact of adding mutations to the analysis.

```{r}
# Read RNAseq
srat = readRDS("data/01_tme_annotation.rds")
srat$cellid = unlist(lapply(str_split(names(srat$orig.ident), "-"), "[[", 1))
```

The table below provides a comprehensive list of all mutations, along with the associated single-cell metadata

```{r}
tmp_genotype = read.csv(genotype_file, row.names = 1)
tmp_genotype$alt <- ifelse(grepl('-', tmp_genotype$alt), '-', tmp_genotype$alt)
tmp_genotype['id'] = paste0(tmp_genotype[,'gene'], tmp_genotype[,'start'], tmp_genotype[,'alt'])
tmp_genotype = tmp_genotype[tmp_genotype$id %in% unique(variants$id),]
tmp_genotype = as_tibble(tmp_genotype[tmp_genotype$gene %in% c("RUNX1", "IDH1", "SRSF2", "FLT3", "CSF3R"),])
tmp_genotype = tmp_genotype[, c("gene", "cell", "genotype")] %>% 
  pivot_wider(names_from = gene, values_from = genotype) %>%
  mutate_all( ~ replace(., . == "", "na"))

metadata = srat@meta.data
metadata = metadata[metadata$sample_id == sample_id, ] 
table2show = merge(tmp_genotype, metadata[,c("cellid", "compartment", "aggregated_ct")], by.x="cell", by.y="cellid")

DT::datatable(
  data.frame(table2show),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

## RUNX1 - IDH1 - FLT3

Describing the mutations in the order RUNX1, IDH1, and FLT3, we observe the following cell counts: 26 cells with mutations in all three genes (MMM), 68 cells with mutations in RUNX1 and IDH1 but not FLT3 (MMW), and 17 cells with mutations only in RUNX1 (MWW).

```{r}
gene_interest <- c('FLT3', 'IDH1', 'RUNX1')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
```



### Effect of FLT3 mutation (MMM vs WMM)

We selected cells where FLT3 is either wild-type or mutated, while both IDH1 and RUNX1 are mutated. This resulted in the following number of cells, representing the intersection of genotyped cells and those passing scRNA-seq filters.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data.  Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
mutation_status1 = c("MUT", "MUT", "MUT")
mutation_status2 = c("WT", "MUT", "MUT")

cells1 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status1, nrow = nrow(subset), ncol = length(mutation_status1), byrow = TRUE), 1, all)]
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```

```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No genes were identified as significantly differentially expressed between the two conditions in the differential expression analysis, likely due to the low number of cells. As a result, we opted to forgo further comparisons, given the limited cell count.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]
```



## RUNX1 - SRSF2 - FLT3

Describing the mutations in the order RUNX1, SRSF2, and FLT3, we observe the following cell counts: 50 cells with mutations in all three genes (MMM), 77 cells with mutations in RUNX1 and IDH1 but not FLT3 (MMW), and 5 cells with mutations only in RUNX1 (MWW).

```{r}
gene_interest <- c('FLT3', 'SRSF2', 'RUNX1')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
```


### Effect of FLT3 mutation (MMM vs WMM)

We selected cells where FLT3 is either wild-type or mutated, while both SRSF2 and RUNX1 are mutated. This resulted in the following number of cells, representing the intersection of genotyped cells and those passing scRNA-seq filters.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
mutation_status1 = c("MUT", "MUT", "MUT")
mutation_status2 = c("WT", "MUT", "MUT")

cells1 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status1, nrow = nrow(subset), ncol = length(mutation_status1), byrow = TRUE), 1, all)]
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```

```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

A single gene 'MGST1' has been identified as significantly differentially expressed between the two conditions in the differential expression analysis, likely due to the low number of cells. As a result, we opted to forgo further comparisons, given the limited cell count.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]
```


# Genotype phenotype analysis: partial genotype

Given the low number of cells, requiring the genotype for all genes was not feasible. Therefore, we relaxed the genotype criteria and assumed that the cells acquire mutations in the following order:

- RUNX1 -> SRSF2 -> FLT3

- RUNX1 -> IDH1 -> FLT3

Therefore, to study the effect of acquiring FLT3 mutations, we can compare FLT3 MUT cells with the combined group of FLT3 WT + IDH1/SRSF2 MUT

## FLT3 MUT vs  FLT3 WT + SRSF2 MUT

We identified 269 cells with FLT3 mutations and 126 cells with FLT3 WT and SRSF2 mutations. The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data.  Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('FLT3')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$FLT3 == "MUT"]

gene_interest <- c('FLT3', 'SRSF2')
mutation_status2 = c("WT", "MUT")

subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

up = reactome_enrichment(markers, down=FALSE)
down = reactome_enrichment(markers, down=TRUE)
```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```

__Enriched in FLT3 Mutated__

```{r}
dotplot(up, title=paste0("Enriched in ", paste(mutation_status1, collapse = "_")), showCategory=15)

DT::datatable(
  up@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)


```

__Enriched in FLT3 WT + SRSF2 Mutated__

```{r}
dotplot(down, title=paste0("Enriched in ", paste(mutation_status2, collapse = "_")), showCategory=15)

DT::datatable(
  down@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```



## FLT3 MUT vs  FLT3 WT + IDH1 MUT

We identified 269 cells with FLT3 mutations and 143 cells with FLT3 WT and IDH1 mutations.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('FLT3')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$FLT3 == "MUT"]

gene_interest <- c('FLT3', 'IDH1')
mutation_status2 = c("WT", "MUT")

subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

up = reactome_enrichment(markers, down=FALSE)
down = reactome_enrichment(markers, down=TRUE)
```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```

__Enriched in FLT3 Mutated__

Overall, these enriched pathways suggest that FLT3-mutated AML cells may exhibit:

- Increased protein synthesis and ribosomal activity to support rapid growth. (e.g., "Peptide chain elongation," "Eukaryotic Translation Initiation")

- Adaptations that mimic viral mechanisms and stress responses, likely aiding in immune evasion and survival. (e.g., "Viral mRNA Translation," "Host Interactions of HIV factors")

0 Metabolic and amino acid adjustments to endure nutrient scarcity and manage oxidative stress. (e.g., "Response of EIF2AK4 (GCN2) to amino acid deficiency," "Metallothioneins bind metals")

- Potential migration and invasion mechanisms that could facilitate interactions with and retention in the bone marrow microenvironment. (e.g., "Regulation of expression of SLITs and ROBOs")

```{r}
dotplot(up, title=paste0("Enriched in ", paste(mutation_status1, collapse = "_")), showCategory=15)

DT::datatable(
  up@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)



```

__Enriched in FLT3 WT + IDH1 Mutated__

```{r}
dotplot(down, title=paste0("Enriched in ", paste(mutation_status2, collapse = "_")), showCategory=15)

DT::datatable(
  down@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```



## FLT3 MUT vs  FLT3 WT 

We identified 269 cells with FLT3 mutations and 610 cells with FLT3 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('FLT3')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$FLT3 == "MUT"]

mutation_status2 = c("WT")
cells2 = subset$cell[subset$FLT3 == "WT"]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

up = reactome_enrichment(markers, down=FALSE)
down = reactome_enrichment(markers, down=TRUE)
```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```

__Enriched in FLT3 Mutated__

```{r}
dotplot(up, title=paste0("Enriched in ", paste(mutation_status1, collapse = "_")), showCategory=15)

DT::datatable(
  up@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)



```

__Enriched in FLT3 WT__

```{r}
dotplot(down, title=paste0("Enriched in ", paste(mutation_status2, collapse = "_")), showCategory=15)

DT::datatable(
  down@result[,c("p.adjust", "Description", "geneID")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```




## IDH1 MUT vs  IDH1 WT 

We identified 438 cells with IDH1 mutations and 148 cells with IDH1 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('IDH1')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$IDH1 == "MUT"]

mutation_status2 = c("WT")
cells2 = subset$cell[subset$IDH1 == "WT"]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No differenetially expressed genes have been detected

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

```


## IDH1 MUT vs  IDH1 WT in FLT3 WT

We identified 143 cells with IDH1 mutations and 33 cells with IDH1 WT in FLT3 WT condition.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('IDH1', 'FLT3')
mutation_status1 = c('MUT', "WT")
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status1, nrow = nrow(subset), ncol = length(mutation_status1), byrow = TRUE), 1, all)]

mutation_status2 = c("WT", "WT")
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No differentially expressed genes have been detected

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

```




## SRSF2 MUT vs  SRSF2 WT 

We identified 459 cells with SRSF2 mutations and 60 cells with IDH1 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated SRSF2 cells
gene_interest <- c('SRSF2')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$SRSF2 == "MUT"]

mutation_status2 = c("WT")
cells2 = subset$cell[subset$SRSF2 == "WT"]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No differenetially expressed genes have been detected

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]
```


## SRSF2 MUT vs  SRSF2 WT in FLT3 WT condition

We identified 126 cells with SRSF2 mutations and 12 cells with IDH1 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
# Extract mutated FLT3 cells
gene_interest <- c('SRSF2', 'FLT3')
mutation_status1 = c('MUT', "WT")
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status1, nrow = nrow(subset), ncol = length(mutation_status1), byrow = TRUE), 1, all)]

mutation_status2 = c("WT", "WT")
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No differentially expressed genes have been detected

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"), 
                       group.by = "mutation_status", 
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]
```

