---
title: "Genotype phenotype AML5"
author: 
- Marcello Del Corvo
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 6,
                      fig.height = 6)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(mclust)
  library(maftools)
  library(ggplot2)
  library(plotly)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(ReactomePA)
  source('src/functions.R')
})

```

# Objective

Genotype data integration of ONT enrichment experiment with 10X expression data for AML5 sample


# Genotype data

The genotype data has been analyzed using [NanoGen](https://github.com/dimadatascience/NanoGen), a custom pipeline designed to call mutations from targeted enrichment data. NanoGen follows a multi-step process to ensure accurate mutation detection. It starts by using Blaze to identify cell barcodes, ensuring that each cell’s data is correctly assigned. It then collapses Unique Molecular Identifiers (UMIs) to eliminate PCR duplicates and other biases, allowing for a more accurate representation of the original sample.

One key feature of NanoGen is its ability to call consensus at the base level, rather than at the sequence level, providing a more precise identification of mutations. NanoGen applies a negative binomial model to distinguish between true mutated reads and background noise, improving the reliability of mutation calls. To address the issue of mutated allele dropout, NanoGen imposes a strict threshold: it requires at least five distinct wild-type (WT) UMIs for a cell to be confidently classified as wild-type (WT).

## Data exploration

First, we identify and retain only mutations that overlap with the WES data. For each candidate mutation, we calculate its cellular fraction. Next, we filter out any genes with fewer than 50 genotyped cells.

```{r}
genotype_file = "data/enrichment/AML5.csv"
wes_file = "data/S34030_AML5_D.small_mutations.cancervar.escat.maf"
sample_id = "AML5"
```

```{r, results=FALSE}
# Read and filter genotyped cells
genotype = read.csv(genotype_file, row.names = 1)
genotype = genotype[genotype$genotype!="", ]
genotype['total_umis'] = (genotype[,'WT'] + genotype[,'MUT'] + genotype[,'MIS'])
genotype$alt <- ifelse(grepl('-', genotype$alt), '-', genotype$alt)
genotype['id'] = paste0(genotype[,'gene'], genotype[,'start'], genotype[,'alt'])

variants = data.frame(read.maf(wes_file)@data)

variants['id'] = paste0(variants[,'Hugo_Symbol'] , variants[,'Start_Position'] , variants[,'Tumor_Seq_Allele2'])
genotype = as_tibble(genotype[genotype$id %in% (variants$id), ])

# Calculate ccf and filter lowly genotyped mutations
ccf = genotype %>%
  group_by(gene, genotype) %>%            # Group by 'gene' and 'genotype'
  summarise(count = n(), .groups = 'drop') %>%  # Count occurrences of each 'genotype'
  pivot_wider(names_from = genotype,
              values_from = count,
              values_fill = 0)  # Pivot to wide format, filling missing values with 0

ccf = ccf[(ccf$MUT + ccf$WT) > 50,]
ccf['ccf'] = ccf[,'MUT'] /(ccf[,'MUT'] + ccf[,'WT']) 

# Add wes
subset <- variants %>%
  filter(id %in% genotype$id) %>%
  select(Hugo_Symbol, tumor_f)

ccf <- ccf %>%
  left_join(subset, by = c("gene" = "Hugo_Symbol"))  %>%
  mutate(wes_ccf = pmin(tumor_f * 2, 1))  # Ensure values > 1 are set to 1

# Calculate correlation
correlation <- cor(ccf$ccf, ccf$wes_ccf, use = "complete.obs")
correlation_title <- sprintf("Correlation: %.2f", correlation)



# Create the scatter plot
p <- ggplot(ccf, aes(x = ccf, y = wes_ccf, text = gene)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  ggtitle(correlation_title) +
  labs(x = "single cell CF", y = "WES_CF") +
  theme(plot.title = element_text(hjust = 0.5, size = 20))

# Convert to interactive plotly plot
fig <- ggplotly(p, tooltip = c("text"))
```

Below, we present the correlation between the cellular fraction, defined as VAF * 2, in WES  and the cellular fraction in the target enrichment. As demonstrated, the target enrichment effectively recapitulates the cellular fraction observed in WES.

```{r}
fig
```


```{r}
DT::datatable(
  ccf,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```


## Mutation co-occurence  

Next, we quantified the number of cells with multiple genotyped mutations, as shown in the tables below. We present the results for different mutation combinations. For this specific samples, we will look at these genes since they are the most genotyped: RUNX1, SRSF2, EZH2

```{r}
# Define the genes of interest
gene_interest <- c('EZH2', 'SRSF2', 'RUNX1')
co_occurence <- get_coccurence(genotype_file, variants, gene_interest)
DT::datatable(
  co_occurence,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)


```

# Genotype phenotype analysis: full genotype

We begin by analyzing the phenotype-genotype relationships using only cells with genotyped data for multiple genes. Specifically, we will focus on the following cases:

1- RUNX1 - SRSF2 - EZH2

The reason is that, for these combination, we have the highest number of genotyped cells for all genes.

For each case, we will examine the impact of adding mutations to the analysis.

```{r}
# Read RNAseq
srat = readRDS("data/01_tme_annotation.rds")
srat$cellid = unlist(lapply(str_split(names(srat$orig.ident), "-"), "[[", 1))
```

The table below provides a comprehensive list of all mutations, along with the associated single-cell metadata

```{r}
tmp_genotype = read.csv(genotype_file, row.names = 1)
tmp_genotype$alt <- ifelse(grepl('-', tmp_genotype$alt), '-', tmp_genotype$alt)
tmp_genotype['id'] = paste0(tmp_genotype[,'gene'], tmp_genotype[,'start'], tmp_genotype[,'alt'])
tmp_genotype = tmp_genotype[tmp_genotype$id %in% unique(variants$id),]
tmp_genotype = as_tibble(tmp_genotype[tmp_genotype$gene %in% c("RUNX1", "SRSF2", "EZH2"),])
tmp_genotype = tmp_genotype[, c("gene", "cell", "genotype")] %>% 
  pivot_wider(names_from = gene, values_from = genotype) %>%
  mutate_all( ~ replace(., . == "", "na"))

metadata = srat@meta.data
metadata = metadata[metadata$sample_id == sample_id, ] 
table2show = merge(tmp_genotype, metadata[,c("cellid", "compartment", "aggregated_ct")], by.x="cell", by.y="cellid")

DT::datatable(
  data.frame(table2show),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```
## RUNX1 - SRSF2 - EZH2

```{r}
gene_interest <- c('EZH2', 'SRSF2', 'RUNX1')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
```

Describing the mutations in the order RUNX1, SRSF2, and EZH2, we observe the following cell counts: `r sum(subset$EZH2=='MUT' & subset$SRSF2=='MUT' & subset$RUNX1=='MUT')` cells with mutations in all three genes (MMM), `r sum(subset$EZH2=='WT' & subset$SRSF2=='MUT' & subset$RUNX1=='MUT')` cells with mutations in RUNX1 and SRSF2 but not EZH2 (MMW), and `r sum(subset$EZH2=='WT' & subset$SRSF2=='WT' & subset$RUNX1=='MUT')` cells with mutations only in RUNX1 (MWW).

### Effect of EZH2 mutation (MMM vs WMM)

We selected cells where EZH2 is either wild-type or mutated, while both SRSF2 and RUNX1 are mutated. This resulted in the following number of cells, representing the intersection of genotyped cells and those passing scRNA-seq filters.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data.  Since the majority of cells belong to the tumor compartment, we focused our analysis solely on the tumor compartment.

```{r}
mutation_status1 = c("MUT", "MUT", "MUT")
mutation_status2 = c("WT", "MUT", "MUT")

cells1 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status1, nrow = nrow(subset), ncol = length(mutation_status1), byrow = TRUE), 1, all)]
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")

```

```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

No genes were identified as significantly differentially expressed between the two conditions in the differential expression analysis, likely due to the low number of cells. As a result, we opted to forgo further comparisons, given the limited cell count.

```{r}
rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"),
                       group.by = "mutation_status",
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]
```
# Genotype phenotype analysis: partial genotype

Given the low number of cells, requiring the genotype for all genes was not feasible. Therefore, we relaxed the genotype criteria and assumed that the cells acquire mutations in the following order:

- RUNX1 -> SRSF2 -> EZH2

Therefore, to study the effect of acquiring EZH2 mutations, we can compare EZH2 MUT cells with the combined group of EZH2 WT + SRSF2 MUT

## EZH2 MUT vs  EZH2 WT + SRSF2 MUT

```{r}
# Extract mutated EZH2 cells
gene_interest <- c('EZH2')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$EZH2 == "MUT"]

gene_interest <- c('EZH2', 'SRSF2')
mutation_status2 = c("WT", "MUT")

subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")

```

We identified `r length(cells1)` cells with EZH2 mutations and `r length(cells2)` cells with EZH2 WT and SRSF2 mutations. The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data.  Since the majority of cells belong to the tumor compartment, we focused our analysis solely on the tumor compartment.


```{r}

DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.
We got only down-regulated genes (genes enriched in EZH2 wild-type + SRSF2 mutated cells), but no pathway was found for this gene-set.

```{r}

rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"),
                       group.by = "mutation_status",
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

down = reactome_enrichment(markers, down=TRUE)

```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```



## EZH2 MUT vs  EZH2 WT

```{r}
# Extract mutated EZH2 cells
gene_interest <- c('EZH2')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$EZH2 == "MUT"]

mutation_status2 = c("WT")
cells2 = subset$cell[subset$EZH2 == "WT"]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```

We identified `r length(cells1)` cells with EZH2 mutations and `r length(cells2)` cells with EZH2 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Againg the majority of cells belong to the undefined compartment and only a small number are from tumor and TME.


```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```
We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.
We got only down-regulated genes (genes enriched in EZH2 wild-type cells), but no pathway was found for this gene-set.

```{r}

rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"),
                       group.by = "mutation_status",
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

down = reactome_enrichment(markers, down=TRUE)

```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```

## SRSF2 MUT vs  SRSF2 WT 

```{r}
# Extract mutated EZH2 cells
gene_interest <- c('SRSF2')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$SRSF2 == "MUT"]

mutation_status2 = c("WT")
cells2 = subset$cell[subset$SRSF2 == "WT"]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")
```

We identified `r length(cells1)` cells with SRSF2 mutations and `r length(cells2)` cells with SRSF2 WT.  The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data. Since the majority of cells belong to the tumor compartment and only a small number are from the TME, we focused our analysis solely on the tumor compartment.

```{r}
DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```

We conducted differential gene expression analysis between the two groups and followed it with reactome pathway enrichment analysis. The dot plot below, along with the accompanying table, presents the enriched terms.
We got only down-regulated genes (genes enriched in SRSF2 wild-type cells), but no pathway was found for this gene-set.

```{r}

rna = subset(rna, subset = compartment =="tumor")
Idents(rna) = rna$mutation_status
markers = FindMarkers(rna, ident.1 = paste(mutation_status1, collapse = "_"), ident.2 = paste(mutation_status2, collapse = "_"),
                       group.by = "mutation_status",
                       min.pct = 0.1, logfc.threshold = 0.25)
markers = markers[markers$p_val_adj < 0.05, ]

down = reactome_enrichment(markers, down=TRUE)

```

__Differential expressed genes__

The table below displays the differentially expressed genes identified in the comparison between `r paste(mutation_status1, collapse = "_")` vs `r paste(mutation_status2, collapse = "_")`


```{r}
DT::datatable(
  markers[,c("avg_log2FC", "p_val_adj")],
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = TRUE
)
```



## SRSF2 MUT vs  SRSF2 WT in EZH2 WT condition


```{r}
# Extract mutated SRSF2 cells
gene_interest <- c('SRSF2')
mutation_status1 = c('MUT')
subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells1 = subset$cell[subset$SRSF2 == "MUT"]

gene_interest <- c('EZH2', 'SRSF2')
mutation_status2 = c("WT", "WT")

subset = get_cellid_full_genotype(genotype_file, variants, gene_interest)
cells2 = subset$cell[apply(subset[,gene_interest] == matrix(mutation_status2, nrow = nrow(subset), ncol = length(mutation_status2), byrow = TRUE), 1, all)]

rna = subset(srat, subset = sample_id == sample_id & cellid %in% c(cells1, cells2))
rna$mutation_status = paste(mutation_status1, collapse = "_")
rna$mutation_status[rna$cellid %in% cells2 ] = paste(mutation_status2, collapse = "_")

```

We identified `r length(cells1)` cells with SRSF2 mutations and `r length(cells2)` cells with EZH2 WT. We got only one relevant tme cell where both SRSF2 and EZH2 are wild-type.
The table below presents the number of cells in each compartment after filtering for cells with available RNA expression data.


```{r}

DT::datatable(
  data.frame(table(rna@meta.data[, c("compartment", "mutation_status")])),
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)
```


