---
title: "Genotype phenotype PALM06"
author: 
- Y Zhan
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 6,
                      fig.height = 6)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(mclust)
  library(maftools)
  library(ggplot2)
  library(plotly)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(ReactomePA)
  source("src/functions.R")
})

```

# Objective

The objective of the analysis is to integrate the genotype data from PALM enrichemnt experiment with 10X expression data.
We start with PALM06.

# Genotype data

The genotype data has been analyzed using [NanoGen](https://github.com/dimadatascience/NanoGen), a custom pipeline designed to call mutations from targeted enrichment data. NanoGen follows a multi-step process to ensure accurate mutation detection. It starts by using Blaze to identify cell barcodes, ensuring that each cell’s data is correctly assigned. It then collapses Unique Molecular Identifiers (UMIs) to eliminate PCR duplicates and other biases, allowing for a more accurate representation of the original sample.

One key feature of NanoGen is its ability to call consensus at the base level, rather than at the sequence level, providing a more precise identification of mutations. NanoGen applies a negative binomial model to distinguish between true mutated reads and background noise, improving the reliability of mutation calls. To address the issue of mutated allele dropout, NanoGen imposes a strict threshold: it requires at least five distinct wild-type (WT) UMIs for a cell to be confidently classified as wild-type (WT).

## Data exploration

For each candidate mutation, we calculate its cellular fraction. Next, we filter out any genes with fewer than 50 genotyped cells.

```{r}
genotype_file = "data/enrichment/PALM06.csv"
wes_file = "data/NUFRA_BM.maf"
sample_ids = c("PALM_06_1", "PALM_06_2")
```



```{r, results=FALSE}
# Read and filter genotyped cells
genotype = read.csv(genotype_file, row.names = 1)
genotype = genotype[genotype$genotype!="", ]
genotype['total_umis'] = (genotype[,'WT'] + genotype[,'MUT'] + genotype[,'MIS'])
genotype$alt <- ifelse(grepl('-', genotype$alt), '-', genotype$alt)
genotype['id'] = paste0(genotype[,'gene'], genotype[,'start'], genotype[,'alt'])

tmp = data.frame(read.maf(wes_file)@maf.silent)
variants = data.frame(read.maf(wes_file)@data)
variants = rbind(variants, tmp)

variants['id'] = paste0(variants[,'Hugo_Symbol'] , variants[,'Start_Position'] , variants[,'Tumor_Seq_Allele2'])
genotype = as_tibble(genotype[genotype$id %in% (variants$id), ])

# Calculate ccf and filter lowly genotyped mutations
ccf = genotype %>%
  group_by(gene, genotype) %>%            # Group by 'gene' and 'genotype'
  summarise(count = n(), .groups = 'drop') %>%  # Count occurrences of each 'genotype'
  pivot_wider(names_from = genotype,
              values_from = count,
              values_fill = 0)  # Pivot to wide format, filling missing values with 0

ccf = ccf[(ccf$MUT + ccf$WT) > 50,]
ccf['ccf'] = ccf[,'MUT'] /(ccf[,'MUT'] + ccf[,'WT']) 

# Add wes
subset <- variants %>%
  filter(id %in% genotype$id) %>%
  select(Hugo_Symbol, tumor_f)

ccf <- ccf %>%
  left_join(subset, by = c("gene" = "Hugo_Symbol"))  %>%
  mutate(wes_ccf = pmin(tumor_f * 2, 1))  # Ensure values > 1 are set to 1

# Calculate correlation
correlation <- cor(ccf$ccf, ccf$wes_ccf, use = "complete.obs")
correlation_title <- sprintf("Correlation: %.2f", correlation)

# Create the scatter plot
p <- ggplot(ccf, aes(x = ccf, y = wes_ccf, text = gene)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  ggtitle(correlation_title) +
  labs(x = "single cell CF", y = "WES_CF") +
  theme(plot.title = element_text(hjust = 0.5, size = 20))

# Convert to interactive plotly plot
fig <- ggplotly(p, tooltip = c("text"))
```

```{r}
fig
```

```{r}
DT::datatable(
  ccf,
  extensions = 'Buttons',
  options = list(
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf'),
    lengthMenu = list(c(10, 25, 50,-1),
                      c(10, 25, 50, "All"))
  ),
  rownames = FALSE
)


```


